Carrier phase and amplitude estimation for phase shift keying using pilots and data

### 2014-05-27 paper.tex
added new data

Figures~\ref{fig:plotphaseBPSK},~\ref{fig:plotphaseQPSK}~and~\ref{fig:plotphase8PSK} show the sample mean square error (MSE) of the phase estimator when $M=2,4,8$ with $L=4096$ and for varying proportions of pilot symbols $\abs{P} = 0, \tfrac{L}{32}, \tfrac{L}{8}, \tfrac{L}{2}, L$.  When $\abs{P} \neq 0$ the mean square error is computed as $\tfrac{1}{T}\sum_{i=1}^T\sfracpart{\hat{\theta}_i - \theta_0}_\pi^2$.  Otherwise, when $\abs{P}=0$ the mean square error is computed as $\tfrac{1}{T}\sum_{i=1}^T\sfracpart{\hat{\theta}_i - \theta_0}^2$ as in~\cite{McKilliam_leastsqPSKnoncoICASSP_2012}.  The dots, squares, circles and crosses are the results of Monte-Carlo simulations with the least square estimator.  The solid lines are the estimator MSEs predicted by Theorem~\ref{thm:normality}.   %(and by Theorem~2 from~\cite{McKilliam_leastsqPSKnoncoICASSP_2012} when $\abs{P} = 0$).  
The prediction is made by dividing the asymptotic covariance matrix by $L$.  The theorem accurately predicts the behaviour of the phase estimator when SNR is sufficiently large.  As the SNR decreases the variance of the phase estimator approaches that of the uniform distribution on $[-\pi, \pi)$ when $\abs{P} \neq 0$, and the uniform distribution on $[-\tfrac{\pi}{M}, \tfrac{\pi}{M})$ when $\abs{P}=0$~\cite{McKilliam_leastsqPSKnoncoICASSP_2012}.  As the SNR increases the variance of the estimators converge to that of the estimator where all symbols are pilots, i.e. $\abs{P} = L$.

@@

Figures~\ref{fig:plotphaseBPSK},~\ref{fig:plotphaseQPSK}~and~\ref{fig:plotphase8PSK} also display the sample MSE of the non-data-aided phase estimator of Viterbi and Viterbi~\cite{ViterbiViterbi_phase_est_1983}.  This estimator requires selection of a function $F$ that transforms the amplitude of each sample prior to the final estimation step.  % (see equation (2) in~\cite{ViterbiViterbi_phase_est_1983}).  
Viterbi and Viterbi propose several viable alternatives, from which we have chosen $F(x) = 1$.  The Viterbi and Viterbi estimator is only applicable in the non-data-aided setting, i.e. when $\abs{P} = 0$.  The sample MSE of the least squares estimator (when $\abs{P} = 0$) and the Viterbi and Viterbi estimator is similar.  The least squares estimator appears slightly more accurate for some values of SNR.
Figures~\ref{fig:plotphaseBPSK},~\ref{fig:plotphaseQPSK}~and~\ref{fig:plotphase8PSK} also display the non-data-aided Cram\'{e}r-Rao lower bound (CRB) for carrier phase given in~\cite{Cowley_crbs_phase_freq_1996,950345}.  These CRBs are derived for the non-data-aided setting and are only valid for estimators that do not make use of pilot symbols, i.e., $\abs{P}=0$.

@@

Figure~\ref{fig:plotampBPSK} %,~\ref{fig:plotampQPSK}~and~\ref{fig:plotamp8PSK} 
shows the mean square error of the amplitude estimator when $M=2$ and $L=2048$ and when the number of pilot symbols is $\abs{P} = 0, \tfrac{L}{2}, L$.  The solid lines show the mean square error predicted by the squared bias $\rho_0^2\big(1 - G(0)\big)^2$ plus the variance of $\hat{m}_L = \hat{\rho} - \rho_0G(0)$ predicted by Theorem~\ref{thm:normality}.  That is, the solid lines are given by
\[
\rho_0^2\big(1 - G(0)\big)^2 + \rho_0^2\frac{p B_1 + d B_2}{L}. 
\]
When SNR is large $G(0)$ is close to $1$ and the bias of the amplitude estimator is small.  However, $G(0)$ grows without bound as the variance of the noise increases so the bias is significant when SNR is small.  Figure~\ref{fig:plotampBPSK} suggests our asymptotic theory to be an excellent predictor of the mean square error of the amplitude estimator.

>>>>

Figures~\ref{fig:plotphaseBPSK},~\ref{fig:plotphaseQPSK}~and~\ref{fig:plotphase8PSK} show the sample mean square error (MSE) of the phase estimator when $M=2,4,8$ with $L=4096$ and for varying proportions of pilot symbols $\abs{P} = 0, \tfrac{L}{32}, \tfrac{L}{8}, \tfrac{L}{2}, L$.  When $\abs{P} \neq 0$ the mean square error is computed as $\tfrac{1}{T}\sum_{i=1}^T\sfracpart{\hat{\theta}_i - \theta_0}_\pi^2$.  Otherwise, when $\abs{P}=0$ the mean square error is computed as $\tfrac{1}{T}\sum_{i=1}^T\sfracpart{\hat{\theta}_i - \theta_0}^2$ as in~\cite{McKilliam_leastsqPSKnoncoICASSP_2012}.  The dots, squares, circles and crosses are the results of Monte-Carlo simulations with the least square estimator.  The solid lines are the estimator MSEs predicted by Theorem~\ref{thm:normality}.   %(and by Theorem~2 from~\cite{McKilliam_leastsqPSKnoncoICASSP_2012} when $\abs{P} = 0$).  
The prediction is made by dividing the asymptotic covariance matrix by $L$.  The theorem accurately predicts the behaviour of the phase estimator when SNR is sufficiently large.  As the SNR decreases the variance of the phase estimator approaches that of the uniform distribution on $[-\pi, \pi)$ when $\abs{P} \neq 0$, and the uniform distribution on $[-\tfrac{\pi}{M}, \tfrac{\pi}{M})$ when $\abs{P}=0$~\cite{McKilliam_leastsqPSKnoncoICASSP_2012}.  As the SNR increases the variance of the estimators converge to that of the estimator where all symbols are pilots, i.e. $\abs{P} = L$.  As expected, the figures indicate the estimator to become more accurate as the number of pilot symbols is increased.  An interesting property distriguishing the case when no pilot symbols exist is the slope of the MSE curve at low SNR.  When a finite proportion of pilot symbols exist the slope at low SNR is similar to that when all symbols are pilots.  The slope is steeper in the case that no pilot symbols exist.

@@

Figures~\ref{fig:plotphaseBPSK},~\ref{fig:plotphaseQPSK}~and~\ref{fig:plotphase8PSK} also display the sample MSE of the non-data-aided phase estimator of Viterbi and Viterbi~\cite{ViterbiViterbi_phase_est_1983}.  This estimator requires selection of a function $F$ that transforms the amplitude of each sample prior to the final estimation step.  % (see equation (2) in~\cite{ViterbiViterbi_phase_est_1983}).  
Viterbi and Viterbi propose several viable alternatives, from which we have chosen $F(x) = 1$.  The Viterbi and Viterbi estimator is only applicable in the non-data-aided setting, i.e., when $\abs{P} = 0$.  The sample MSE of the least squares estimator (when $\abs{P} = 0$) and the Viterbi and Viterbi estimator is similar.  The least squares estimator appears slightly more accurate for some values of SNR.
Figures~\ref{fig:plotphaseBPSK},~\ref{fig:plotphaseQPSK}~and~\ref{fig:plotphase8PSK} also display the non-data-aided Cram\'{e}r-Rao lower bound (CRB) for carrier phase given in~\cite{Cowley_crbs_phase_freq_1996,950345}.  These CRBs are derived for the non-data-aided setting and are only valid for estimators that do not make use of pilot symbols, i.e., $\abs{P}=0$.

@@

Figures~\ref{fig:plotampBPSK},~\ref{fig:plotampQPSK}~and~\ref{fig:plotamp8PSK} show the mean square error of the amplitude estimator when $M=2,4,8$ and $L=4096$ and when the number of pilot symbols is $\abs{P} = 0, \tfrac{L}{2}, L$.  The solid lines show the mean square error predicted by the squared bias $\rho_0^2\big(1 - G(0)\big)^2$ plus the variance of $\hat{m}_L = \hat{\rho} - \rho_0G(0)$ predicted by Theorem~\ref{thm:normality}.  That is, the solid lines are given by
\[
\rho_0^2\big(1 - G(0)\big)^2 + \rho_0^2\frac{p B_1 + d B_2}{L}. 
\]
When the SNR is large $G(0)$ is close to $1$ and the bias of the amplitude estimator is small.  However, $G(0)$ grows without bound as the variance of the noise increases so the bias is significant when SNR is small.  Figure~\ref{fig:plotampBPSK},~\ref{fig:plotampQPSK}~and~\ref{fig:plotamp8PSK} suggests our asymptotic theory to be an excellent predictor of the mean square error of the amplitude estimator.

### 2014-05-27 response2.tex
added new data

\item \textbf{Comment:} p. 3: ``all of the exiting bounds are applicable only in the fully non-data-aided case or in the completely data aided case \dots'' --> bounds for the common scenario where there is a mixture of pilot and data symbols can be found in "N. Noels, H. Steendam, M. Moeneclaey en H. Bruneel, Carrier Phase and Frequency Estimation for Pilot Symbol Assisted Transmission: Bounds and Algorithms, IEEE TSP, Vol. 53, No. 12, p. 4578-4587, Dec. 2005. \\
\textbf{Response:} We have added this reference and have removed the sentence begining ``To the authors' knowledge, all of the existing bounds are applicable only in the fully non-data-aided case \dots''. Note that the model considered in the paper by Noels~et.~al. treats the amplitude as fixed and known and includes a frequency offset term.  Thus, the bounds are not immediately applicable in our setting and we are not able to meaningfully compare these bounds with our simulation results.  For this reason with have kept with the bound described in~\cite{950345} as in the previous version of the manuscript.

@@

\item \textbf{Comment:} in general, I find the numerical results section rather short \\
\textbf{Response:} We have greatly expanded the simulation section. BLERG: this includes \dots ref to below.

\item \textbf{Comment:} although different figures are shown for M=2,4,8 (phase estimation), the effect of the constellation size is not addressed in the text  \\
\textbf{Response:}

@@

\item \textbf{Comment:}  although the most important contribution of the paper is with respect to the biased amplitude estimate, only 1 figure is presented for amplitude estimation, namely for $L=2048$ and BPSK $\to$ this seems not well balanced  \\
\textbf{Response:} Figures have been added showing the amplitude estimation error for BSPK, QPSK, and 8-PSK. BLERG: Add to text!

\item \textbf{Comment:}  a separate plot of the two terms that contribute to Fig. 4 would be welcome \\
\textbf{Response:}

>>>>

\item \textbf{Comment:} p. 3: ``all of the exiting bounds are applicable only in the fully non-data-aided case or in the completely data aided case \dots'' --> bounds for the common scenario where there is a mixture of pilot and data symbols can be found in "N. Noels, H. Steendam, M. Moeneclaey en H. Bruneel, Carrier Phase and Frequency Estimation for Pilot Symbol Assisted Transmission: Bounds and Algorithms, IEEE TSP, Vol. 53, No. 12, p. 4578-4587, Dec. 2005. \\
\textbf{Response:} We have added this reference and have removed the sentence begining ``To the authors' knowledge, all of the existing bounds are applicable only in the fully non-data-aided case \dots''. Note that the model considered in the paper by Noels~et.~al.~\cite{Noels_crbs_with_pilots_2005} treats the amplitude as fixed and known and includes a frequency offset term.  Thus, the bounds are not immediately applicable in our setting and we are not able to meaningfully compare these bounds with our simulation results.  %For this reason with have kept with the bound described in~\cite{950345} as in the previous version of the manuscript.

@@

\item \textbf{Comment:} in general, I find the numerical results section rather short \\
\textbf{Response:} We have greatly expanded the simulation section as detailed in the responses that follow.

\item \textbf{Comment:} although different figures are shown for M=2,4,8 (phase estimation), the effect of the constellation size is not addressed in the text  \\
\textbf{Response:}  We have included the follows sentences to Section~\ref{sec:simulations}:

``As expected, the figures indicate the estimator to become more accurate as the number of pilot symbols is increased.  An interesting property distriguishing the case when no pilot symbols exist is the slope of the MSE curve at low SNR.  When a finite proportion of pilot symbols exist the slope at low SNR is similar to that when all symbols are pilots.  The slope is steeper in the case that no pilot symbols exist.''

@@

\item \textbf{Comment:}  although the most important contribution of the paper is with respect to the biased amplitude estimate, only 1 figure is presented for amplitude estimation, namely for $L=2048$ and BPSK $\to$ this seems not well balanced  \\
\textbf{Response:} Figures have been added showing the amplitude estimation error for BSPK, QPSK, and 8-PSK.

\item \textbf{Comment:}  a separate plot of the two terms that contribute to Fig. 4 would be welcome \\

\textbf{Response:}
